---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE
)
```

# routing

<!-- badges: start -->
<!-- badges: end -->


`routing` is an R port of [pillarjs/router](https://github.com/pillarjs/router), the routing mechanism that underpins Express.js. It gives R web frameworks a syntax that mimics how Express.js applications are written, so the same patterns that make Express.js applications easy to read and reason about are now available in R.


## Installation
 
```{r, eval=FALSE}
remotes::install_github("JulioCollazos64/routing")
```

Let's look at an example:

```{r}
library(routing)
router <- Router$new()
router$use(function(req, res) {
  print(req)
})
router$get("/hello", function(req, res) {
  res$send("Hello world!")
})
```

Even if we knew nothing about web development, we can clearly see the things that matter from this code snippet:

- **Router** - We create an object called `router` that holds all the rules for how incoming requests should be handled.
- **Middleware** - `router$use()` appends a piece of code that runs on every requests, regardless of path or method.
- **Route definition**
  - **HTTP verb** - `get` means this route only responds to HTTP GET requests
  - **PATH** - `/hello` is the URL path this route matches

Moreover, the routing mechanism happens in the *same order you write your code* so you can read how your request goes from top to bottom.

## Usage in Web Frameworks:

You can get pretty far by using only `routing` for your web development projects. In the
following example, we have a router with only one route and two handlers, half of the time 
a **call** to this **app** returns "HEADS" and the other half returns "TAILS" depending on
the value of `runif(1)`

```{r}
library(routing)
router <- Router$new()

router$get(
  "/",
  function(req, res) {
    if (runif(1) < .5) {
      return(forward())
    }

    list(
      status = 200L,
      headers = list(
        "Content-Type" = "text/html"
      ),
      body = "HEADS"
    )
  },
  function(req, res) {
    list(
      status = 200L,
      headers = list(
        "Content-Type" = "text/html"
      ),
      body = "TAILS"
    )
  }
)


responses <- vector(mode = "character", length = 1000)
for (i in seq_len(1000)) {
  request <- fiery::fake_request("http://your-next-project.com/")
  response <- router$handle(request)
  responses[i] <- response$body
}


table(responses)
```

`routing` is meant to be used in pair with `httpuv` as such we can translate
our boilerplate code into a real application with minimal effort:

```{r}
server <- httpuv::startServer(
  "0.0.0.0",
  8080,
  list(
    call = function(req) {
      router$handle(req)
    }
  )
)
```

Now visit <http://localhost:8080/> and see your web application in action. That's all
it takes to include to include `routing` in your R web framework! Now you have
request parameters support through [pater](https://github.com/JulioCollazos64/pater) and all
the features of Express routing.

### Create a Response Class

However, as you may have noticed already we had to manually specify our server response
using `list()`, which is error-prone and time consuming. Moreover, we lack common
needs such as response serialization to json.

We need a better way of dealing with this. That's why it's advisable to create a response
object at the start of your app call, here we provide a minimal response class but you can
easily imagine one with more features. 


```{r}
Response <- R6::R6Class(
  "Response",
  public = list(
    status = NULL,
    headers = list(),
    send = function(body = NULL) {
      list(
        status = self$status,
        headers = self$headers,
        body = body
      )
    }
  )
)
```

In order to pass this object through the router stack we will assign it to our
Router handle method: `router$handle(req, Response$new())`

Our handlers now can call the send method from the response class, saving a few
keystrokes.

### Use the default "catch" everything handler

If in our application we were to visit http://localhost:8080/api we won't get
any response back from the server as we didn't define a route for that path. For
this cases it's better to play safe and return a generic response saying
that resource in our server couldn't be found. As it's such a common need
`routing` provides this fallback automatically through the `finalHandler` function
which catches any unhandled request.


Out final application would look like:

```{r, eval=FALSE}
server <- httpuv::startServer(
  "0.0.0.0",
  8080,
  list(
    call = function(req) {
      res <- Response$new()
      router$handle(req, res, finalHandler(req, res))
    }
  )
)
```


## How it differs from Express.js

1. We use `forward()` instead of `next()` to pass control to the next handler.
2. You don't need to declare `forward` as an argument to your handler to call it inside.
3. If your handler doesn't return a response or call `forward()`, one is called on your behalf.