---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE
)
```

# routing

<!-- badges: start -->
<!-- badges: end -->


`routing` is an R port of [pillarjs/router](https://github.com/pillarjs/router), the routing mechanism that underpins Express.js. It gives R web frameworks a syntax that mimics how Express.js applications are written, so the same patterns that make Express.js applications easy to read and reason about are now available in R.


## Installation
 
```{r, eval=FALSE}
remotes::install_github("JulioCollazos64/routing")
```

Let's look at an example:

```{r}
library(routing)
router <- Router$new()
router$use(function(req, res) {
  print(req)
})
router$get("/hello", function(req, res) {
  res$send("Hello world!")
})
```

Even if we knew nothing about web development, we can clearly see the things that matter from this code snippet:

- **Router** - We create an object called `router` that holds all the rules for how incoming requests should be handled.
- **Middleware** - `router$use()` appends a piece of code that runs on every requests, regardless of path or method.
- **Route definition**
  - **HTTP verb** - `get` means this route only responds to HTTP GET requests
  - **PATH** - `/hello` is the URL path this route matches

Moreover, the routing mechanism happens in the *same order you write your code* so you can read how your request goes from top to bottom.

## Usage in Web Frameworks:

You can get pretty far by using only `routing` for your web development projects. In the
following example, we have a router with only one route and two handlers, half of the time 
a **call** to this **app** returns "HEADS" and the other half returns "TAILS" depending on
the value of `runif(1)`

```{r}
library(routing)
router <- Router$new()

router$get(
  "/",
  function(req, res) {
    if (runif(1) < .5) {
      return(forward())
    }

    list(
      status = 200L,
      headers = list(
        "Content-Type" = "text/html"
      ),
      body = "HEADS"
    )
  },
  function(req, res) {
    list(
      status = 200L,
      headers = list(
        "Content-Type" = "text/html"
      ),
      body = "TAILS"
    )
  }
)


responses <- vector(mode = "character", length = 1000)
for (i in seq_len(1000)) {
  request <- fiery::fake_request("http://your-next-project.com/")
  response <- router$handle(request)
  responses[i] <- response$body
}


table(responses)
```

`routing` is meant to be used in pair with `httpuv` as such we can translate
our boilerplate code into a real application with minimal effort:

```{r}
server <- httpuv::startServer(
  "0.0.0.0",
  8080,
  list(
    call = function(req) {
      router$handle(req)
    }
  )
)
```

Now visit <http://localhost:8080/> and see your web application in action. That's all
it takes to include to include `routing` in your R web framework! Now you have
request parameters support through [pater](https://github.com/JulioCollazos64/pater) and all
the features of Express routing.

### Create a Response Class

However, as you may have noticed already we had to manually specify our server response
using `list()`, which is error-prone and time consuming. Moreover, we lack common
needs such as response serialization to json.

We need a better way of dealing with this. That's why it's advisable to create a response
object at the start of your app call, here we provide a minimal response class but you can
easily imagine one with more features. 


```{r}
Response <- R6::R6Class(
  "Response",
  public = list(
    status = NULL,
    headers = list(),
    send = function(body = NULL) {
      list(
        status = self$status,
        headers = self$headers,
        body = body
      )
    }
  )
)
```

In order to pass this object through the router stack we will assign it to our
Router handle method: `router$handle(req, Response$new())`

Our handlers now can call the send method from the response class, saving a few
keystrokes.

### Use the default "catch" everything handler

If in our application we were to visit http://localhost:8080/api we won't get
any response back from the server as we didn't define a route for that path. For
this cases it's better to play safe and return a generic response saying
that resource in our server couldn't be found. As it's such a common need
`routing` provides this fallback automatically through the `finalHandler` function
which catches any unhandled request.


Out final application would look like:

```{r, eval=FALSE}
server <- httpuv::startServer(
  "0.0.0.0",
  8080,
  list(
    call = function(req) {
      res <- Response$new()
      router$handle(req, res, finalHandler(req, res))
    }
  )
)
```

## How a request goes through the Layer stack

A Router in `routing` is a stack of routes, middlewares or other routers, conceptually
each of them are a *Layer* of the stack, when a request comes in each Layer is
checked *in the order they were added to the stack*, no ordering occurs. 
Whether a Layer matches or not a request depends not only on its current **matcher**, a 
function that determines whether theirs a mtach between the request PATH and the Layer path,
but also on the previous Layers as **a Layer can modify the request PATH**. 

```{r}
router <- Router$new()
router$get(
  "/",
  function(req, res) {
    writeLines(
      c("Handler for", req$PATH_INFO)
    )

    req$PATH_INFO <- "/foo"
  }
)

router$get("/foo", function(req, res) {
  list(
    status = 200L,
    body = "foo",
    headers = list(
      "Content-Type" = "text/html"
    )
  )
})

req <- fiery::fake_request("http://your-next-project/")

router$handle(req)
```

As you can see from this example  the first route acted only as a middleware
for *url-rewriting*, that's why we say whether a Layer matches a given request depends on the
previous Layers too. 

### A Layer matcher

A Layer matcher is built with the help of `match` from  the [pater](https://cran.r-project.org/web/packages/pater/index.html) package, 
it uses the path you passed to `$use()` or `$route` (or its HTTP verb shortcuts), contrary
to Ambiorix this path isn't appended to the "Router basepath", let's look at an example:

```{r}
app <- Router$new()
api <- Router$new()

api$get(c("/users", "/usuarios"), function(req, res) {
  list(
    status = 200L,
    body = "We've got many users!",
    headers = list(
      "Content-Type" = "text/html"
    )
  )
})

app$use("/api", api)

req <- fiery::fake_request("http://your-next-project.com/api/users")
app$handle(req)
```

We said previously that a Router can have another as a Layer, and each Layer have
it's own matcher based on the path you've actually written. So `app` Router has a 
Router `api` with a matcher for any request starting with "/api"

Our route for "/users" and "/usuarios" as a Layer has its own matcher set to match
any request that matches those paths, in case you're wondering here are the actual
regexes used to match the request PATH.

```{r}
lapply(
  api$getStack()[[1]]$matchers,
  FUN = function(s) {
    environment(s)$regex
  }
)
```

You can see from those regexes that neither of them actually matches the path
"/api/users", what happens internally to make these sort of regexes actually
match is *strip/trim* the basePath of the Router to the request path, so that it 
goes from "/api/users" to "/users", which actually matches.

### Layers everywhere

The design in Express routing is such that the concept of Layer is also used
in the Route class, this make it possible write code like the following:

```{r}
router <- Router$new()
logger <- function(req, res) {
  print(req$PATH_INFO)
}

router$get(
  "/foo",
  # route specific middleware
  logger,
  # route handler
  function(req, res) {
    list(
      status = 200L,
      body = "bar",
      headers = list(
        "Content-Type" = "text/html"
      )
    )
  },
  function(req, res) {
    print("Never called")
  }
)

layer <- router$getStack()[[1]]
# A route's Layer stack
layer$route$stack
```

So when a request comes to "/foo" it matches the first Router Layer, which happens to
be a Route, then the route dispatch the request into its own stack of layers, which in turn 
process the request until one of them actually respond, in this case this happens at the 
second Layer.

In summary, when we think of a Route Layer we see that it has its own way of 
handling the request instead of just being the `handlers` we just wrote it's a method
that dispatches the requests into their handlers. Just like a Router dispatches a
request into a middleware, route or another Router.



## How it differs from Express.js

1. We use `forward()` instead of `next()` to pass control to the next handler.
2. You don't need to declare `forward` as an argument to your handler to call it inside.
3. If your handler doesn't return a response or call `forward()`, one is called on your behalf.